\chapter{Position and Momentum}
\section{Introduction}
So far, we have considered {\em general} dynamical variables represented by 
{\em general} linear
operators acting in ket space. However, in classical mechanics the most important
dynamical variables are those involving
position and momentum. Let us investigate 
the role of such  variables  in quantum mechanics. 

In classical mechanics, the position $q$ and momentum $p$ 
of some component of a dynamical  system are represented as {\em real numbers} which,
by definition, commute.
In quantum mechanics, these quantities are represented
as   {\em noncommuting} linear Hermitian operators acting in a ket space
which represents all of the possible states of the system. Our first task is
to discover a quantum mechanical replacement for the classical result
$q\,p-p\,q = 0$. Do the position and momentum operators commute? If not,
what is the value of $q\,p-p\,q$? 

\section{Poisson Brackets}
Consider a dynamic system whose state at a particular time $t$ is
fully specified 
by $N$ independent classical  coordinates $q_i$ (where $i$ runs from 1 to $N$). 
Associated with each generalized coordinate $q_i$ is a 
classical canonical momentum
$p_i$. For instance, a Cartesian coordinate  has an associated linear
momentum, an angular coordinate has an associated angular momentum, {\em etc}.
As is well-known, the behaviour of a classical system can be specified in terms
of Lagrangian or Hamiltonian dynamics. For instance, in Hamiltonian dynamics,
\begin{eqnarray}
\frac{d q_i}{d t} &=& ~\frac{\partial H}{\partial p_i},\\[0.5ex]
\frac{d p_i}{dt} &=& - \frac{\partial H}{\partial q_i},
\end{eqnarray}
where the function $H(q_i, p_i, t)$ is the energy of the system at time $t$
expressed in terms of the 
classical coordinates and canonical momenta. This function is
usually referred to as the {\em Hamiltonian} of the system.

We are interested in 
finding some
construct of classical dynamics which consists of
{\em products} of dynamical variables. If such a construct exists we hope to 
generalize it somehow to obtain  a
rule describing how dynamical variables
commute with one another in quantum mechanics. There is, indeed,
 one well-known construct
in classical dynamics which involves products of dynamical variables. The
{\em Poisson bracket} of two dynamical variables $u$ and $v$ is defined
\begin{equation}\label{e3.2}
[u, v] = \sum_{i=1}^N \left(\frac{\partial u}{\partial q_i}\frac{\partial v}
{\partial p_i} - \frac{\partial u}{\partial p_i}\frac{\partial v}{\partial q_i}
\right),
\end{equation}
where $u$ and $v$ are regarded as functions of the coordinates 
and momenta $q_i$ and $p_i$. It is easily demonstrated that
\begin{eqnarray}\label{e3.3a}
[q_i, q_j] &=& 0,\\[0.5ex]
[p_i, p_j] &=& 0,\\[0.5ex]
[q_i, p_j] &=& \delta_{ij}.\label{e3.3c}
\end{eqnarray}
 The time evolution of a dynamical variable can also
be
written in terms of a Poisson bracket by noting that
\begin{eqnarray}
\frac{du}{dt}& = &\sum_{i=1}^N \left(\frac{\partial u}{\partial q_i}\frac{d q_i}{dt}
+ \frac{\partial u}{\partial p_i}\frac{dp_i}{dt}\right)\nonumber\\[0.5ex]
&=& \sum_{i=1}^N \left(\frac{\partial u}{\partial q_i}\frac{\partial H}{\partial p_i}
-\frac{\partial u}{\partial p_i}\frac{\partial H}{\partial q_i}\right)\nonumber
\\[0.5ex]
&=& [u, H],\label{e3.4}
\end{eqnarray}
where use has been made of Hamilton's equations. 

Can we construct a quantum mechanical Poisson bracket in which $u$ and
$v$ are noncommuting operators, instead of functions? Well, the main properties
of the classical Poisson bracket are as follows:
\begin{eqnarray}\label{e3.5a}
[u, v] &=& - [v, u],\\[0.5ex]
[u, c] &=& 0,\\[0.5ex]
[u_1+ u_2, v] &=& [u_1, v] + [u_2, v],\\[0.5ex]
[u, v_1 + v_2]&=& [u, v_1] + [u, v_2]\\[0.5ex]
[u_1\, u_2, v] &=& [u_1, v] u_2 + u_1 [u_2, v],\label{e3.5e}\\[0.5ex]
[u, v_1 \,v_2] &=& [u, v_1] v_2 + v_1 [u, v_2],\label{e3.5f}
\end{eqnarray}
and
\begin{equation}\label{e3.6}
[u, [v, w] ]+ [v, [w, u] ] + [w, [u, v]] = 0.
\end{equation}
The last relation is known as the {\em Jacobi identity}. In the above,
$u$, $v$, $w$, {\em etc.}, represent dynamical variables, and $c$ represents a number.
Can we find some combination of noncommuting operators $u$ and $v$, {\em etc.}, 
which satisfies all of the above relations?

Well, we can evaluate the Poisson bracket $[u_1 \,u_2, v_1 \,v_2]$ in
two different ways, since we can use either of the formulae (\ref{e3.5e}) or
(\ref{e3.5f}) first. Thus,
\begin{eqnarray}
[u_1\, u_2, v_1\, v_2] &=& [u_1, v_1 \,v_2]u_2 + u_1[u_2, v_1\, v_2]\\[0.5ex]
&=& \left\{ [u_1, v_1]v_2 + v_1[u_1, v_2]\right\} u_2 
+u_1\left\{[u_2, v_1]v_2 + v_1[u_2, v_2]\right\}\nonumber\\[0.5ex]
&=& [u_1, v_1] v_2 \,u_2 + v_1[u_1, v_2] u_2 + u_1[u_2, v_1]v_2
+ u_1 \,v_1[u_2, v_2],\nonumber
\end{eqnarray}
and
\begin{eqnarray}
[u_1 \,u_2, v_1\, v_2] &=& [u_1 \,u_2, v_1 ]v_2 + v_1[u_1\, u_2, v_2]\\[0.5ex]
&=& [u_1, v_1] u_2 \,v_2 + u_1[u_2, v_1] v_2 + v_1[u_1, v_2]u_2
+ v_1\, u_1[u_2, v_2].\nonumber
\end{eqnarray}
Note that the order of the various factors has been preserved, since they
now represent {\em noncommuting} operators. Equating the above two results
yields
\begin{equation}
[u_1, v_1] (u_2 \,v_2 - v_2 \,u_2) = (u_1 \,v_1-v_1\, u_1)[u_2, v_2].
\end{equation}
Since this relation must hold for $u_1$ and $v_1$ quite independent of
$u_2$ and $v_2$, it follows that
\begin{eqnarray}
u_1 \,v_1 - v_1\, u_1 &=& {\rm i}\,\hbar \,[u_1, v_1],\\[0.5ex]
u_2\, v_2 - v_2\, u_2 &=& {\rm i} \,\hbar \,[u_2, v_2],
\end{eqnarray}
where $\hbar$ does not depend on $u_1$, $v_1$, $u_2$, $v_2$, and also
commutes with $(u_1\, v_1- v_1 \,u_1)$. Since $u_1$, {\em etc.}, are quite general
operators, it follows that $\hbar$ is just a number. We want the quantum 
mechanical Poisson
bracket of two Hermitian operators to be an Hermitian operator itself, since
the classical Poisson bracket of two real dynamical variables is real. This
requirement is satisfied if $\hbar$ is a real number. Thus, the
quantum mechanical Poisson bracket of two dynamical variables $u$ and $v$
is given by
\begin{equation}
[u, v] = \frac{u\,v - v\,u}{{\rm i} \,\hbar},
\end{equation}
where $\hbar$ is a new universal constant of nature. Quantum mechanics agrees with
experiments provided that  $\hbar$ takes the value $h/2\pi$, where
\begin{equation}
h = 6.6261 \times 10^{-34}~~{\rm J\,s}
\end{equation}
is {\em Planck's constant}. Somewhat confusingly, the notation
$[u, v]$ is
conventionally  reserved for the commutator $u\,v-v\,u$ in quantum mechanics. 
We will use $[u, v]_{\rm quantum}$ to denote the quantum Poisson bracket. 
Thus,
\begin{equation}\label{e3.13}
[u, v]_{\rm quantum} = \frac{[u, v]}{{\rm i}\, \hbar}.
\end{equation}
It is easily demonstrated that the quantum mechanical Poisson bracket, as defined above,
satisfies all of the relations (\ref{e3.5a})--(\ref{e3.6}).

The strong analogy we have found between the classical Poisson bracket, defined
in Eq.~(\ref{e3.2}), and the quantum mechanical 
Poisson bracket, defined in Eq.~(\ref{e3.13}), leads
us to make the assumption that the quantum mechanical bracket has the same
value as the corresponding classical bracket, at least for the simplest
cases. In other words, we are assuming that Eqs.~(\ref{e3.3a})--(\ref{e3.3c}) hold for quantum
mechanical as well as classical Poisson brackets. This argument yields the
fundamental commutation relations
\begin{eqnarray}
[q_i, q_j] &=& 0,\label{e3.14a}\\[0.5ex]
[p_i, p_j]&=& 0, \label{e3.14b}\\[0.5ex]
[q_i, p_j] &=& {\rm i} \,\hbar \,\delta_{ij}.\label{e3.14c}
\end{eqnarray}
These results provide us with the basis for calculating commutation
relations between general dynamical variables. For instance, if
two dynamical variables,  $\xi$ and $\eta$,
can both  be written as a power series in the
 $q_i$ and $p_i$, then repeated application of Eqs.~(\ref{e3.5a})--(\ref{e3.5f})
allows $[\xi, \eta]$ to be expressed in terms of the fundamental
commutation relations (\ref{e3.14a})--(\ref{e3.14c}). 

Equations~(\ref{e3.14a})--(\ref{e3.14c}) provide the foundation for the analogy between quantum mechanics
and classical mechanics. Note that the classical result (that everything commutes)
is obtained in the limit $\hbar\rightarrow 0$. Thus, {\em classical mechanics
can be regarded as the limiting case of quantum mechanics when
$\hbar$ goes to zero}. 
In classical mechanics, each 
pair of generalized coordinate and its conjugate momentum, $q_i$ and
$p_i$, correspond to a different classical degree of freedom of the system. 
It is clear from Eqs.~(\ref{e3.14a})--(\ref{e3.14c}) that in quantum mechanics the {\em dynamical
variables corresponding to different degrees of freedom all commute}. 
It is only those variables corresponding to the same degree of freedom which
may fail to commute.

\section{Wavefunctions}
Consider a simple system with one classical degree of freedom, which corresponds to
the Cartesian coordinate $x$. Suppose that $x$ is free to take any value ({\em e.g.},
$x$ could be the position of a free particle). The classical dynamical variable
$x$ is represented in quantum 
mechanics  as a linear Hermitian operator which is also called  $x$.
Moreover, the operator $x$ possesses eigenvalues $x'$
lying in the {\em continuous}
range $-\infty< x'<+\infty$ (since the eigenvalues
correspond to all the possible results of a measurement of $x$). We can
span ket space using the suitably normalized eigenkets of $x$.
An eigenket corresponding to the eigenvalue $x'$ is denoted $|x'\rangle$.
Moreover, 
[see Eq.~(\ref{e2.77})]
\begin{equation}\label{e3.15}
\langle x' | x''\rangle = \delta(x'-x'').
\end{equation}
The eigenkets satisfy the extremely useful relation [see Eq.~(\ref{e2.78})]
\begin{equation}\label{e3.16}
\int_{-\infty}^{+\infty} d x' \, |x'\rangle\langle x'|= 1.
\end{equation}
This formula expresses the fact that the eigenkets are complete, mutually
orthogonal, and suitably normalized.

A state ket $|A\rangle$ (which represents a general  state $A$ of the system)
can be expressed as a linear superposition of the eigenkets of the position
operator using Eq.~(\ref{e3.16}). Thus,
\begin{equation}\label{e3.17}
|A\rangle = \int_{-\infty}^{+\infty} dx' \langle x'|A\rangle |x'\rangle
\end{equation}
The quantity $\langle x'|A\rangle$ is a complex function of the position eigenvalue
$x'$. We can write
\begin{equation}\label{e3.18}
\langle x'|A\rangle = \psi_A(x').
\end{equation}
Here, $\psi_A(x')$ is the famous {\em wave-function} of quantum mechanics. 
Note that state $A$ is completely specified by its wave-function $\psi_A(x')$
[since the wave-function can be used to reconstruct the state ket $|A\rangle$
using Eq.~(\ref{e3.17})].
It is clear that the wave-function of state $A$ is simply the collection
of the weights of the corresponding state ket $|A\rangle$,
when it is expanded in terms of the eigenkets of the
position operator. Recall, from Sect.~\ref{s2.10}, that the probability of
a measurement of a dynamical variable $\xi$ yielding the result $\xi'$ 
when the system is in state $A$ is given by
$|\langle \xi'|A\rangle|^2$, assuming that
the
eigenvalues of $\xi$ are discrete. This result is easily generalized to dynamical
variables possessing continuous eigenvalues. In fact, the probability of
a measurement of $x$ yielding a result lying in the range $x'$ to $x'+dx'$
when the system is in a state $|A\rangle$ is $|\langle x'|A\rangle|^2\,dx'$. 
In other words, the probability of a measurement of position yielding a
result in the range $x'$ to $x'+dx'$ when the wave-function of the system is
$\psi_A(x')$ is
\begin{equation}\label{e3.19}
P(x', dx') = |\psi_A(x')|^2\, dx'.
\end{equation}
This formula is only valid if the state ket $|A\rangle$ is properly normalized:
{\em i.e.}, if $\langle A|A\rangle = 1$. The corresponding normalization for
the wave-function is
\begin{equation}\label{e3.20}
\int_{-\infty}^{+\infty} |\psi_A(x')|^2\,dx'= 1.
\end{equation}
Consider a second state $B$ represented by a state ket $|B\rangle$ and
a wave-function $\psi_B(x')$. The inner product $\langle B| A\rangle$ 
can be written
\begin{equation}
\langle B| A\rangle = \int_{-\infty}^{+\infty} dx'\,\langle B| x'\rangle
\langle x' | A \rangle 
 = \int_{-\infty}^{+\infty} \psi_B^\ast (x') \,\psi_A'(x')\,dx',
\end{equation}
where use has been made of Eqs.~(\ref{e3.16}) and (\ref{e3.18}). Thus, the inner product of two states is
related to the overlap integral of their wave-functions. 

Consider a general function $f(x)$ of the observable $x$ [{\em e.g.}, $f(x)=x^2$]. 
If $|B\rangle = f(x)|A\rangle$ then it follows that
\begin{eqnarray}
\psi_B(x')& = &\langle x'| f(x) \int_{-\infty}^{+\infty} dx''\,\psi_A(x'')|x''\rangle
\nonumber\\[0.5ex]
&=& \int_{-\infty}^{+\infty} dx''\, f(x'')\,\psi_A(x'') \langle x'|x''\rangle,
\end{eqnarray}
giving
\begin{equation}
\psi_B(x') = f(x')\, \psi_A(x'),
\end{equation}
where use has been made of Eq.~(\ref{e3.15}). Here, $f(x')$ is the same function
of the position eigenvalue $x'$ that $f(x)$ is of the position operator $x$:
{\em i.e.}, if $f(x)=x^2$ then $f(x') = x'^{\,2}$. It follows, from the above result,
that a general state ket $|A \rangle$ can be written
\begin{equation}
|A\rangle = \psi_A(x) \rangle,
\end{equation}
where $\psi_A(x)$ is the same function of the operator $x$ that the wave-function
$\psi_A(x')$ is of the position eigenvalue $x'$, and the ket $\rangle$ has the
wave-function $\psi(x') =1$. The ket $\rangle$ is termed the {\em standard ket}. 
The dual of the standard ket is termed the {\em standard bra}, and is
denoted $\langle$.  It is
easily seen that
\begin{equation}\label{e3.25}
\langle \psi_A^\ast(x) \stackrel{\rm DC}{\longleftrightarrow} \psi_A(x)\rangle.
\end{equation}
Note, finally, that $\psi_A(x)\rangle$ is often shortened to $\psi_A\rangle$, leaving
the dependence on the position operator $x$ tacitly understood. 

\section{Schr\"{o}dinger's Representation - I}\label{s3.4}
Consider the simple system described in the previous section. A general
state ket can be written $\psi(x)\rangle$, where $\psi(x)$ is a general function of
the position operator $x$, and $\psi(x')$ is the associated wave-function. 
Consider the ket whose wave-function is $d\psi(x')/dx'$. This ket is
denoted $d\psi/dx\rangle$. The new ket is clearly a linear function of
the  original ket, so we can think of it as the result of some linear
operator acting on $\psi\rangle$. Let us denote this operator
$d/dx$. It follows that
\begin{equation}
\frac{d}{dx} \psi\rangle = \frac{d\psi}{dx}\rangle.
\end{equation}

Any linear operator which acts on ket vectors can also act on bra vectors. 
Consider $d/dx$ acting on a general bra $\langle \phi(x)$. According to
Eq.~(\ref{e2.32}), the bra $\langle \phi \,d/dx$ satisfies
\begin{equation}
\left( \langle \phi \frac{d}{dx} \right) \psi\rangle = \langle \phi\left(
\frac{d}{dx}\psi\rangle\right).
\end{equation}
Making use of Eqs.~(\ref{e3.16}) and (\ref{e3.18}), we can write
\begin{equation}
\int_{-\infty}^{+\infty} \langle \phi \frac{d}{dx} |x'\rangle
\,dx'\, \psi(x') = \int_{-\infty}^{+\infty} \phi(x')\, dx'\, \frac{d\psi(x')}{dx'}.
\end{equation}
The right-hand side can be transformed via integration by parts to give
\begin{equation}\label{e3.29}
\int_{-\infty}^{+\infty} \langle \phi \frac{d}{dx} |x'\rangle
\,dx' \,\psi(x') = -\int_{-\infty}^{+\infty} \frac{d \phi(x')}{dx'}\,  dx' \,\psi(x'),
\end{equation}
assuming that the contributions from the limits of integration vanish. 
It follows that
\begin{equation}
\langle \phi \frac{d}{dx} | x'\rangle = - \frac{d \phi(x')}{dx'},
\end{equation}
which implies 
\begin{equation}\label{e3.31}
\langle \phi \frac{d}{dx} = - \langle \frac{d\phi}{dx}.
\end{equation}
The neglect of contributions from the limits of integration in Eq.~(\ref{e3.29}) is
reasonable because
 physical wave-functions are square-integrable [see Eq.~(\ref{e3.20})].
Note that
\begin{equation}
\frac{d}{dx} \psi\rangle = \frac{d\psi}{dx}\rangle \stackrel{\rm DC}{\longleftrightarrow}
\langle \frac{d\psi^\ast}{dx} = -\langle \psi^\ast \frac{d}{dx},
\end{equation}
where use has been made of Eq.~(\ref{e3.31}). 
It follows, by comparison with Eqs.~(\ref{e2.33}) and (\ref{e3.25}), that
\begin{equation}\label{e3.33}
\left(\frac{d}{dx}\right)^{\dag }= - \frac{d}{dx}.
\end{equation}
Thus, $d/dx$ is an {\em anti-Hermitian}  operator. 

Let us evaluate the commutation relation between the operators $x$ and $d/dx$.
We have
\begin{equation}
\frac{d}{dx} x\, \psi\rangle = \frac{d (x\,\psi)}{dx}\rangle = x\,\frac{d}{dx}\psi\rangle
+ \psi\rangle.
\end{equation}
Since this holds for any ket $\psi\rangle$, it follows that
\begin{equation}\label{e3.35}
\frac{d}{dx} x - x \frac{d}{dx} = 1.
\end{equation}
Let $p$ be the momentum conjugate to $x$
 (for the simple system under consideration
 $p$ is a straight-forward linear momentum). According to Eq.~(\ref{e3.14c}),
$x$ and $p$ satisfy the commutation relation
\begin{equation}\label{e3.36}
x\,p - p\,x = {\rm i} \,\hbar.
\end{equation}
It can be seen, by comparison with Eq.~(\ref{e3.35}), that 
the Hermitian operator {\em $ -{\rm i}\,\hbar \,d/dx$
satisfies the same commutation relation with $x$ that $p$ does}. 
The most general conclusion which may be drawn from a comparison of Eqs.~(\ref{e3.35})
and (\ref{e3.36}) is that
\begin{equation}\label{e3.37}
p = -{\rm i}\,\hbar\,\frac{d}{dx} + f(x),
\end{equation}
since (as is easily demonstrated) a general function $f(x)$ of the position
operator  automatically commutes with $x$. 

We have chosen to normalize the eigenkets and
eigenbras of the position operator so that they satisfy
the normalization condition (\ref{e3.15}). However, this choice of normalization does not
uniquely determine the eigenkets and
eigenbras. Suppose that we transform to a new
set of eigenbras which are related to the old set via
\begin{equation}\label{e3.38}
\langle x'|_{\rm new} = {\rm e}^{\,{\rm i}\,\gamma'}\langle x'|_{\rm old},
\end{equation}
where $\gamma'\equiv \gamma(x')$ is a real function of $x'$.
This transformation amounts to a rearrangement of the relative
phases of the eigenbras.  The new
normalization condition is
\begin{eqnarray}
\langle x'|x''\rangle_{\rm new} &=& \langle x'| {\rm e}^{\,{\rm i}\,\gamma'}
 {\rm e}^{-{\rm i}\,\gamma''}|x''\rangle_{\rm old} 
= {\rm e}^{\,{\rm i}\,(\gamma'-\gamma'')}
\langle x'|x''\rangle_{\rm old}\nonumber\\[0.5ex]
 &=&{ \rm e}^{\,{\rm i}\,(\gamma'-\gamma'')}\delta(x'-x'')= \delta(x'-x'').
\end{eqnarray}
Thus, the new eigenbras satisfy the same normalization condition
 as the old eigenbras. 


By definition, the standard ket $\rangle$ satisfies $\langle x'|\rangle = 1$.
It follows from Eq.~(\ref{e3.38}) that the new standard ket is related to the
old standard ket via
\begin{equation}\label{e3.40}
\rangle_{\rm new} = {\rm e}^{-{\rm i}\, \gamma} \rangle_{\rm old},
\end{equation}
where $\gamma\equiv \gamma(x)$ is a real function of the position operator $x$. 
The dual of the above equation yields the transformation rule for
the standard bra,
\begin{equation}\label{e3.41}
\langle_{\rm new} = \langle_{\rm old} \,{\rm e}^{\,{\rm i}\, \gamma}.
\end{equation}
The transformation rule for a general operator $A$ follows from Eqs.~(\ref{e3.40}) and (\ref{e3.41}),
plus the requirement that the triple product $\langle A\rangle$ remain
invariant (this must be the case, otherwise the probability of a measurement
yielding a certain result would depend on the choice of eigenbras). Thus,
\begin{equation}\label{e3.42}
A_{\rm new} = {\rm e}^{-{\rm i}\,\gamma} A_{\rm old}\,{\rm e}^{\,{\rm i}\,\gamma}.
\end{equation}
Of course, if $A$ commutes with $x$ then $A$ is invariant under the transformation.
In fact, $d/dx$ is the only operator (we know of) which does not commute
with $x$, so Eq.~(\ref{e3.42}) yields
\begin{equation}
\left(\frac{d}{dx}\right)_{\rm new} =  {\rm e}^{-{\rm i}\,\gamma}\frac{d}{dx}\,{\rm e}^{\,{\rm i}\,\gamma} = \frac{d}{dx} + {\rm i}\, \frac{d\gamma}{dx},
\end{equation}
where the subscript ``old'' is taken as read. It follows, from Eq.~(\ref{e3.37}), that
the momentum operator $p$ can be written
\begin{equation}
p = -{\rm i} \,\hbar\,\left(\frac{d}{dx}\right)_{\rm new} - \hbar\,\frac{d\gamma}{dx} + f(x).
\end{equation}
Thus, the special choice
\begin{equation}\label{e3.45}
\hbar\,\gamma(x) = \int^x f(x)\,dx
\end{equation}
 yields
\begin{equation}\label{e3.46}
p =  -{\rm i} \,\hbar\,\left(\frac{d}{dx}\right)_{\rm new}.
\end{equation}
Equation~(\ref{e3.45}) fixes $\gamma$ to
within an arbitrary additive constant: {\em i.e.}, the special eigenkets and eigenbras 
for which Eq.~(\ref{e3.46}) is true are determined to within an arbitrary common phase-factor. 

In conclusion, it is possible to find a set of basis eigenkets and eigenbras
of the position operator $x$ which satisfy the normalization condition (\ref{e3.15}),
and for which the momentum conjugate to $x$ can be represented as the operator
\begin{equation}
p = -{\rm i}\,\hbar\,\frac{d}{dx}.
\end{equation}
A general state ket is written $\psi(x)\rangle$, where the standard ket $\rangle$
satisfies $\langle x'|\rangle = 1$, and where $\psi(x')= \langle x'|
\psi(x)\rangle$ is the wave-function. 
This scheme of things is known as {\em Schr\"{o}dinger's representation}, and is the
basis of wave mechanics. 

\section{Schr\"{o}dinger's Representation - II}\label{s3.5}
In the preceding sections, we have developed Schr\"{o}dinger's representation
for the case of a single 
operator $x$ corresponding to a classical Cartesian coordinate. However, this scheme
can easily be extended. Consider a system with $N$ generalized coordinates, 
$q_1\cdots q_N$, which can all be {\em simultaneously}\/ measured. These are represented 
as $N$ {\em  commuting}\/ operators, $q_1\cdots q_N$, each with a continuous range
of eigenvalues, $q_1'\cdots q_N'$.
 Ket space is conveniently spanned  by  the simultaneous
eigenkets of $q_1\cdots q_N$, which are denoted $|q_1'\cdots q_N'\rangle$. These 
eigenkets must form a complete set, otherwise the $q_1\cdots q_N$ would not be 
simultaneously observable. 

The orthogonality condition for the eigenkets [{\em i.e.}, the generalization of
Eq.~(\ref{e3.15})] is
\begin{equation}
\langle q_1'\cdots q_N'| q_1''\cdots q_N''\rangle = \delta(q_1'-q_1'')\,\delta(q_2'-q_2'')\cdots
\delta(q_N'-q_N'').
\end{equation}
The completeness condition [{\em i.e.}, the generalization of Eq.~(\ref{e3.16})] is
\begin{equation}
\int_{-\infty}^{+\infty} \cdots\int_{-\infty}^{+\infty} dq_1' \cdots dq_N'\,
|q_1'\cdots q_N'\rangle \langle q_1'\cdots q_N'| = 1.
\end{equation}
The standard ket $\rangle$ is defined such that
\begin{equation}\label{e3.50}
\langle q_1'\cdots q_N'|\rangle = 1.
\end{equation}
The standard bra $\langle$ is the dual of the standard ket. A general state 
ket is written
\begin{equation}
\psi(q_1\cdots q_N)\rangle.
\end{equation}
The associated wave-function is
\begin{equation}
\psi(q_1'\cdots q_N') = \langle q_1'\cdots q_N'|\psi\rangle.
\end{equation}
Likewise, a general state bra is written
\begin{equation}
\langle \phi(q_1\cdots q_N),
\end{equation}
where
\begin{equation}
\phi(q_1'\cdots q_N') = \langle \phi|q_1'\cdots q_N'\rangle.
\end{equation}
The probability of an observation of the system finding the first coordinate in
the range $q_1'$ to $q_1'+dq_1'$, the second coordinate in the range $q_2'$ to $q_2'
+dq_2'$, {\em etc.}, is
\begin{equation}
P(q_1'\cdots q_N'; dq_1'\cdots dq_N') = |\psi(q_1'\cdots q_N')|^2\,dq_1'\cdots 
dq_N'.
\end{equation}
Finally, the normalization condition for a physical wave-function is
\begin{equation}
\int_{-\infty}^{+\infty} \cdots \int_{-\infty}^{+\infty}
|\psi(q_1'\cdots q_N')|^2\,dq_1'\cdots dq_N' = 1.
\end{equation}

The $N$ linear operators $\partial/\partial q_i$ (where $i$ runs from 1 to $N$)
are defined
\begin{equation}\label{e3.57}
\frac{\partial}{\partial q_i} \psi\rangle =\frac{\partial\psi}{\partial q_i}
\rangle.
\end{equation}
These linear operators can also act on bras (provided the associated wave-functions
are square integrable) in accordance with [see Eq.~(\ref{e3.31})]
\begin{equation}
\langle \phi \frac{\partial }{\partial q_i} = -\langle \frac{\partial \phi}
{\partial q_i}.
\end{equation}
Corresponding to Eq.~(\ref{e3.35}), we can derive the commutation relations
\begin{equation}
\frac{\partial}{\partial q_i} q_j - q_j \frac{\partial}{\partial q_i} = \delta_{ij}.
\end{equation}
It is also clear that
\begin{equation}
\frac{\partial}{\partial q_i}\frac{\partial}{\partial q_j} \psi\rangle
= \frac{\partial^2\psi}{\partial q_i \partial q_j} \rangle = \frac{\partial }{\partial
q_j} \frac{\partial }{\partial q_i} \psi\rangle,
\end{equation}
showing that
\begin{equation}
\frac{\partial}{\partial q_i}\frac{\partial}{\partial q_j} = 
\frac{\partial }{\partial q_j}
\frac{\partial }{\partial q_i}.
\end{equation}

It can be seen, by comparison with Eqs.~(\ref{e3.14a})--(\ref{e3.14c}), that the linear operators
$-{\rm i}\,\hbar\, \partial/\partial q_i$ satisfy the same commutation relations
with the $q$'s
and with each other that the $p$'s do. The most general conclusion
we can draw from this coincidence of commutation relations is (see Dirac)
\begin{equation}
p_i = -{\rm i} \,\hbar \frac{\partial}{\partial q_i} +
\frac{\partial  F(q_1\cdots q_N) }{\partial q_i}.
\end{equation}
However, the function $F$
 can be transformed away via  a suitable readjustment of the phases
of the basis eigenkets (see Sect.~\ref{s3.4}, and Dirac). Thus, we can always construct
a set of simultaneous eigenkets of $q_1\cdots q_N$ for  which
\begin{equation}\label{e3.63}
p_i = -{\rm i}\,\hbar \frac{\partial}{\partial q_i}.
\end{equation}
This is the generalized Schr\"{o}dinger representation. 

It follows from Eqs.~(\ref{e3.50}), (\ref{e3.57}), and (\ref{e3.63}) that
\begin{equation}
p_i \rangle = 0.
\end{equation}
Thus, the standard ket in Schr\"{o}dinger's representation is a simultaneous eigenket
of all the momentum operators belonging to the eigenvalue zero. Note that
\begin{equation}
\langle q_1'\cdots q_N'|\frac{\partial}{\partial q_i} \psi\rangle = 
\langle q_1'\cdots q_N'| \frac{\partial \psi}{\partial q_i}\rangle 
=\frac{\partial \psi(q_1'\cdots q_N')}{\partial q_i'} = \frac{\partial }
{\partial q_i'}\langle q_1'\cdots q_N'|\psi\rangle.
\end{equation}
Hence,
\begin{equation}
\langle q_1'\cdots q_N'| \frac{\partial}{\partial q_i} = 
\frac{\partial}{\partial q_i'} \langle q_1'\cdots q_N'|,
\end{equation}
so that
\begin{equation}\label{e3.67}
\langle q_1'\cdots q_N'| p_i = -{\rm i}\,\hbar\frac{\partial}{\partial q_i'}
\langle q_1'\cdots q_N'|.
\end{equation}
The dual of the above equation gives
\begin{equation}
p_i|q_1'\cdots q_N'\rangle = {\rm i}\,\hbar\frac{\partial}{\partial q_i'} |q_1'\cdots
q_N'\rangle.
\end{equation}

\section{Momentum Representation}
Consider a system with one degree of freedom, describable in terms of a coordinate
$x$ and its conjugate momentum $p$, both of which have a continuous range of
eigenvalues. We have seen that it is possible to represent the system in terms
of the eigenkets of $x$. 
This is termed Schr\"{o}dinger's representation.
However, it is also possible to represent the system in 
terms of the eigenkets of $p$. 

Consider the eigenkets of $p$ which belong to the eigenvalues $p'$. These are
denoted $|p'\rangle$. The orthogonality relation for the momentum eigenkets is
\begin{equation}\label{e3.69}
\langle p'|p''\rangle = \delta(p'-p''),
\end{equation}
and the corresponding completeness relation is
\begin{equation}\label{e3.70}
\int_{-\infty}^{+\infty} dp' |p'\rangle \langle p'| = 1.
\end{equation}
A general state ket can be written
\begin{equation}
\phi(p)\rangle
\end{equation}
where the standard ket $\rangle $  satisfies 
\begin{equation}
\langle p'| \rangle = 1.
\end{equation}
Note that the standard ket in this representation is quite
different to that  in Schr\"{o}dinger's representation. 
The momentum space wave-function $\phi(p')$ satisfies
\begin{equation}
\phi(p') = \langle p'|\phi \rangle.
\end{equation}
The probability that a measurement of the momentum yields a result
lying  in the range $p'$ to
$p'+dp'$ is given by
\begin{equation}
P(p', dp') = |\phi(p')|^2\,dp'.
\end{equation}
Finally, the normalization condition for a physical momentum space wave-function is
\begin{equation}
\int_{-\infty}^{+\infty} |\phi(p')|^2\,dp' = 1.
\end{equation}

The fundamental commutation relations (\ref{e3.14a})--(\ref{e3.14c}) exhibit a 
particular symmetry between coordinates and their conjugate momenta. If all the
coordinates are transformed into their conjugate momenta, and {\em vice versa}, and
$\rm i$ is  then replaced by $-\rm i$,  the commutation relations are
unchanged. It follows from this symmetry that we can always choose the eigenkets
of $p$ in such a manner that  the coordinate 
$x$ can be represented as (see Sect.~\ref{s3.4})
\begin{equation}
x = {\rm i}\,\hbar\, \frac{d}{dp}.
\end{equation}
This is termed the momentum representation. 

The above result is easily generalized to a system with more than one degree of
freedom. Suppose the system is specified by $N$ coordinates, $q_1\cdots q_N$, and
$N$ conjugate momenta, $p_1\cdots p_N$. Then, in the momentum representation, the
coordinates can be written as 
\begin{equation}
q_i = {\rm i}\,\hbar \frac{\partial}{\partial p_i}.
\end{equation}
We also have
\begin{equation}
q_i \rangle = 0,
\end{equation}
and
\begin{equation}
\langle p_1' \cdots p_N'| q_i = {\rm i}\,\hbar \frac{\partial}{\partial p_i'}
\langle p_1'\cdots p_N'|.
\end{equation}

The momentum representation is less useful than Schr\"{o}dinger's representation
for  a very simple reason. The energy operator ({\em i.e.}, the Hamiltonian) of
most simple systems takes the form of a sum of quadratic terms in the momenta
({\em i.e.}, the kinetic energy) plus a complicated function of the coordinates
({\em i.e.}, the potential energy). In Schr\"{o}dinger's representation,
the eigenvalue problem for the energy translates into a second-order differential
equation in the coordinates, with a complicated potential function. In the
momentum representation, the problem transforms into a high-order differential
equation in the momenta, with a quadratic potential. With the mathematical
tools at our disposal, we are far better able to solve the former type of problem
than the latter. Hence, Schr\"{o}dinger's representation is
generally more useful than the momentum representation. 

\section{Uncertainty Relation}
How is a momentum space wave-function related to the corresponding coordinate
space wave-function? To answer this question, let us consider the
representative $\langle x'|p'\rangle$ of the
 momentum eigenkets $|p'\rangle$ in Schr\"{o}dinger's representation
for a system with a single degree of freedom. This representative satisfies
\begin{equation}
p' \langle x'|p'\rangle = \langle x'|p|p'\rangle = -{\rm i}\,\hbar\frac{d}{dx'}
\langle x'|p'\rangle,
\end{equation}
where use has been made of Eq.~(\ref{e3.67}) (for the case of a system with one
degree of freedom). The solution of the above  differential equation is
\begin{equation}
\langle x' | p'\rangle = c' \exp({\rm i}\, p' x'/\hbar),
\end{equation}
where $c' = c'(p')$. It is easily demonstrated that
\begin{equation}
\langle p'|p''\rangle = \int_{-\infty}^{+\infty} \langle p'|x'\rangle \,dx'\,
\langle x'|p''\rangle = c'^\ast c'' \int_{-\infty}^{\infty}\exp[-{\rm i}\,
(p'-p'')\,x'/\hbar]\,dx'.
\end{equation}
The well-known mathematical result
\begin{equation}\label{e3.83}
\int_{-\infty}^{+\infty} \exp({\rm i}\, a\,x)\,dx = 2\pi\,\delta (a),
\end{equation}
yields
\begin{equation}
\langle p'|p''\rangle = |c'|^2 \, h\, \delta(p'-p'').
\end{equation}
This is consistent with Eq.~(\ref{e3.69}), provided that $c' = h^{-1/2}$. Thus,
\begin{equation}\label{e3.85}
\langle x'|p'\rangle = h^{-1/2} \exp({\rm i}\, p'\, x'/\hbar).
\end{equation}

Consider a general state ket $|A\rangle$ whose coordinate wave-function is $\psi(x')$,
and whose momentum wave-function is $\Psi(p')$. In other words,
\begin{eqnarray}
\psi(x') &= &\langle x'|A\rangle, \\[0.5ex]
\Psi(p') &=& \langle p'|A\rangle.
\end{eqnarray}
It is easily demonstrated that
\begin{eqnarray}
\psi(x') &=& \int_{-\infty}^{+\infty} dp' \langle x'|p'\rangle \langle p'|A\rangle
\nonumber\\[0.5ex]
&= &\frac{1}{h^{1/2}} \int_{-\infty}^{+\infty} \Psi(p') 
\exp({\rm i} \,p'\,x'/\hbar)\,dp'\label{e3.87}
\end{eqnarray}
and 
\begin{eqnarray}
\Psi(p') &=& \int_{-\infty}^{+\infty} dx' \langle p'|x'\rangle \langle x'|A\rangle
\nonumber\\[0.5ex]
&=& \frac{1}{h^{1/2}} \int_{-\infty}^{+\infty} \psi(x') \exp(-{\rm i}\, p' x'/\hbar)\,
dx',
\end{eqnarray}
where use has been made of Eqs.~(\ref{e3.16}), (\ref{e3.70}), (\ref{e3.83}), and (\ref{e3.85}). 
Clearly, the momentum
space wave-function is the {\em Fourier transform}
 of the coordinate space wave-function. 

Consider a state whose coordinate space wave-function is a {\em wave-packet}. 
In other words, the wave-function only has non-negligible amplitude in some
spatially localized region of extent $\Delta x$. As is well-know, the Fourier
transform of a wave-packet fills up a wave-number band of approximate extent
$\delta k \sim 1/\Delta x$. Note that in Eq.~(\ref{e3.87}) the role of the wave-number
$k$ is played by the quantity $p'/\hbar$. It follows that the momentum space
wave-function corresponding to a wave-packet in coordinate space extends over
a range of momenta $\Delta p \sim \hbar /\Delta x$. Clearly, a measurement 
of $x$ is almost certain to give a result lying in a
range of width $\Delta x$. Likewise, measurement of $p$ is almost certain to
yield a result lying in a range of width $\Delta p$. The product of these two
uncertainties is 
\begin{equation}
\Delta x \,\Delta p \sim \hbar.
\end{equation}
This result is called {\em Heisenberg's uncertainty principle}. 

Actually, it is possible to write Heisenberg's uncertainty principle
 more exactly by making use of
Eq.~(\ref{e2.75}) and the commutation relation (\ref{e3.36}). We obtain
\begin{equation}
\langle (\Delta x)^2\rangle \langle (\Delta p)^2\rangle \geq \frac{\hbar^2}{4}
\end{equation}
for any general state. It is easily demonstrated that the minimum uncertainty states,
for which the equality sign holds in the above relation, correspond to Gaussian
wave-packets in both coordinate and momentum space. 

\section{Displacement Operators}\label{s3.8}
Consider a system with one degree of freedom corresponding to the Cartesian
coordinate $x$. Suppose that we displace this system some distance along the $x$-axis.
We could imagine that the system is on wheels, and we just give it a little
push. The final state of the system is completely determined by its initial state,
together with the direction and magnitude of the displacement.
Note that the type of displacement we are considering is one in which
{\em everything} to do with the system is displaced. So, if the system is
subject to an external potential, then the potential must be  displaced.

The situation is not so clear with state kets. The final state
of the system only determines the {\em direction} of the displaced state ket. Even if
we adopt the  convention that all state kets have unit norms, the final ket is
still not completely determined, since it can be multiplied by a constant phase-factor. However, we know that the superposition relations between states 
remain invariant under the displacement. This follows because the superposition
relations have a physical significance which is unaffected by a displacement of
the system. 
Thus, if 
\begin{equation}\label{e3.91}
|R\rangle = |A\rangle + |B\rangle
\end{equation}
in the undisplaced system, and the displacement causes ket $|R\rangle$ to
transform to ket $|Rd\rangle$, {\em etc}., then in the displaced system we have
\begin{equation}\label{e3.92}
|Rd\rangle = |Ad\rangle + |Bd\rangle.
\end{equation}
Incidentally, this determines the displaced kets to within a single arbitrary phase-factor to be multiplied into all of them. The displaced kets cannot be multiplied by 
individual phase-factors, because this would wreck the superposition relations. 

Since Eq.~(\ref{e3.92}) holds in the displaced system whenever Eq.~(\ref{e3.91}) holds in the
undisplaced system, it follows that the displaced ket $|Rd\rangle$ must be the
result of some linear operator acting on the undisplaced ket
$|R\rangle$. In other
words,
\begin{equation}
|R d\rangle = D |R\rangle,
\end{equation}
where $D$ an  operator which depends only on the
nature of the displacement. The arbitrary phase-factor by which all
displaced kets may be multiplied results in $D$ being undetermined to an arbitrary
multiplicative constant of modulus unity. 

We now adopt the ansatz that any combination of bras, kets, and dynamical
variables which possesses a physical significance is invariant under a displacement
of the system. The normalization condition 
\begin{equation}
\langle A|A\rangle = 1
\end{equation}
for a state ket $|A\rangle$ certainly has a physical significance. Thus, we must
have 
\begin{equation}
\langle Ad|Ad\rangle = 1.
\end{equation}
Now, $|Ad\rangle = D|A\rangle$ and $\langle Ad| = \langle A|D^{\dag}$, so
\begin{equation}
\langle A| D^{\dag} \,D|A\rangle = 1.
\end{equation}
Since this must hold for any state ket $|A\rangle$, it follows that
\begin{equation}\label{e3.97}
D^{\dag} \,D = 1.
\end{equation}
Hence, the displacement operator is {\em unitary}. 
Note that the above relation implies that
\begin{equation}
|A\rangle = D^{\dag} |A d\rangle.
\end{equation}

The  equation
\begin{equation}
v \,|A\rangle = |B\rangle,
\end{equation}
where the operator
$v$ represents a dynamical variable, has some physical significance. Thus,
we require that
\begin{equation}
v_d |Ad\rangle = |Bd\rangle,
\end{equation}
where $v_d$ is the displaced operator. It follows that
\begin{equation}
v_d |Ad\rangle = D |B\rangle = D \,v \,|A\rangle = D\, v\, D^{\dag} |Ad\rangle.
\end{equation}
Since this is true for any ket $|Ad\rangle$, we have
\begin{equation}\label{e3.102}
v_d = D\, v\, D^{\dag}.
\end{equation}
Note that the arbitrary numerical factor in $D$ does not affect either of the
results (\ref{e3.97}) and (\ref{e3.102}).

Suppose, now, that the system is displaced an {\em infinitesimal} distance $\delta x$
along the $x$-axis. We expect that the displaced ket $|A d\rangle$ should
approach the undisplaced ket 
$|A\rangle$ in the limit as $\delta x\rightarrow 0$. Thus,
we expect the limit
\begin{equation}
\lim_{\delta x\rightarrow 0 } \frac{|A d\rangle - |A\rangle}{\delta x}
= \lim_{\delta x\rightarrow 0 }\frac{D-1}{\delta x}|A\rangle
\end{equation}
to exist. Let
\begin{equation}
d_x =  \lim_{\delta x\rightarrow 0 }\frac{D-1}{\delta x},
\end{equation}
where $d_x$ is denoted the {\em displacement operator} along the $x$-axis. The fact 
that $D$ can be replaced by $D \exp({\rm i}\,\gamma)$, where $\gamma$ is a real
phase-angle, implies that $d_x$ can be replaced by
\begin{equation}
 \lim_{\delta x\rightarrow 0 }\frac{D\exp({\rm i}\,\gamma)-1}{\delta x}=
 \lim_{\delta x\rightarrow 0 }\frac{D-1+{\rm i}\, \gamma}{\delta x}= d_x + {\rm i}\,a_x,
\end{equation}
where $a_x$ is the limit of $\gamma/\delta x$. We have assumed, as seems
reasonable,  that $\gamma$ tends
to zero as $\delta x\rightarrow 0$. It is clear that the displacement operator
is undetermined to an arbitrary imaginary additive constant. 

For small $\delta x$, we have
\begin{equation}\label{e3.106}
D = 1 + \delta x\,d_x.
\end{equation}
It follows from Eq.~(\ref{e3.97}) that
\begin{equation}
(1+ \delta x\,d_x^{~\dag}) (1+  \delta x\,d_x) = 1.
\end{equation}
Neglecting order $(\delta x)^2$, we obtain
\begin{equation}
d_x^{~\dag} + d_x = 0.
\end{equation}
Thus, the displacement operator is {\em anti-Hermitian}. Substituting into
Eq.~(\ref{e3.102}), and again neglecting order $(\delta x)^2$, we find that
\begin{equation}
v_d = (1+ \delta x\,d_x)\, v\, (1- \delta x\,d_x) = v + \delta x\,( d_x\, v - v\, d_x),
\end{equation}
which implies
\begin{equation}\label{e3.110}
\lim_{\delta x\rightarrow 0} \frac{v_d -v}{\delta x} = d_x \,v -v\, d_x.
\end{equation}

Let us consider a specific example. 
Suppose that a state has a wave-function $\psi(x')$. If the system is displaced
a distance $\delta x$ along the $x$-axis then the new wave-function is
$\psi(x'-\delta x)$ ({\em i.e.}, the same shape shifted in the $x$-direction
by a distance $\delta x$). Actually, the new wave-function can be multiplied by
an arbitrary number of modulus unity. It can be seen that the new wave-function  
is obtained from the old wave-function according to the 
prescription  $x'\rightarrow x'- \delta x$. Thus,
\begin{equation}
x_d = x -\delta x.
\end{equation}
A comparison  with Eq.~(\ref{e3.110}), using  $x=v$, yields
\begin{equation}
d_x \,x - x\,d_x = -1.
\end{equation}
It follows that ${\rm i}\,\hbar\, d_x$ obeys the same commutation relation with 
$x$ that $p_x$, the momentum conjugate to $x$, does [see Eq.~(\ref{e3.14c})]. 
The most general conclusion we can draw from this observation is that
\begin{equation}
p_x = {\rm i}\,\hbar\, d_x + f(x),
\end{equation}
where $f$ is Hermitian (since $p_x$ is Hermitian). However, the fact that $d_x$
is undetermined to an arbitrary additive imaginary constant (which could be a function
of $x$) enables us to transform the function $f$ out of the above equation, leaving
\begin{equation}\label{e3.114}
p_x = {\rm i}\,\hbar\, d_x.
\end{equation}
Thus, the displacement operator in the $x$-direction is proportional to the
momentum conjugate to $x$. We say that $p_x$ is the {\em generator} of translations
along the $x$-axis. 

A finite translation along the $x$-axis can be constructed from 
a series of very many
infinitesimal translations. Thus, the operator $D(\Delta x)$ which translates the
system a distance $\Delta x$ along the $x$-axis is
written
\begin{equation}
D(\Delta x) =
\lim_{N\rightarrow \infty}
 \left(1-{\rm i}\, \frac{\Delta x}{N} \frac{p_x}{\hbar}\right)^N,
\end{equation}
where use has been made of Eqs.~(\ref{e3.106}) and (\ref{e3.114}). It follows that
\begin{equation}\label{e3.116}
D(\Delta x) = \exp\left({-\rm i} \,p_x\,\Delta x /\hbar\right).
\end{equation}
The unitary nature of the operator is now clearly apparent.

We can also construct displacement operators which translate the system along
the $y$- and $z$-axes. Note that a displacement a distance $\Delta x$ 
along the
$x$-axis {\em commutes} with a displacement a distance $\Delta y$ along the
 $y$-axis. 
In other words, if the system is moved $\Delta x$ along the $x$-axis, and then
$\Delta y$ along the $y$-axis then it ends up in the same state as if it were moved
$\Delta y$ along the $y$-axis, and then $\Delta x$ along the $x$-axis. The fact that
translations in independent directions commute is clearly associated with the
fact that the conjugate momentum operators
 associated with these directions also commute
[see Eqs.~(\ref{e3.14b}) and (\ref{e3.116})].
 